ABM simulating short-term mating behaviors

### 2022-06-10 submission/manuscript.Rmd
final edits

\vspace{30mm}

Yurun Ying^1^^,^^$\dagger$^, Jan Antfolk^2^, Pekka Santtila^1^^,^^3^

^1^ Faculty of Arts and Sciences, NYU Shanghai

^2^ Faculty of Arts, Psychology and Theology, Åbo Akademi University

^3^ NYU-ECNU Institute for Social Development at NYU Shanghai

@@

\vspace{40mm}

^$\dagger$^ To whom correspondence should be addressed: [yurun.ying@nyu.edu](yurun.ying@nyu.edu)

@@

Gender differences in short-term mating behaviors are well-documented in human sexuality research. Existing studies usually conflate gender differences in mating preferences with differences in sexual behaviors, which is theoretically problematic. Using an agent-based model, we investigated the circumstances under which men and women’s differential preferences for short-term mating would result in gender differences in short-term mating behaviors. The model showed that when all individuals in a closed heterosexual population were considered,  men and women had the same average number of short-term mating experiences and short-term mates even when men had stronger preferences for short-term mating. Men (vs. women) had a higher average number of both experiences and mates when analyses were limited to only heterosexual men and women who successfully participated in the mating pool (i.e., those with a non-zero number of short-term mating experiences). Moreover, when men (vs. women) had stronger preferences for short-term mating, gay men had a higher average number of both experiences and mates compared to both lesbian women and heterosexual men. These results suggest that even when gender differences in mating preferences exist, those in short-term mating behaviors only occur among particular populations, or when men’s preferences for short-term mating are not constrained by those of women. Suggestions for future research in human mating psychology and behaviors were provided.

@@

As for mating standards, men are less selective in accepting someone as a potential short-term mate. For example, studies using U.S. college samples have found that the minimum percentile ranks of the desirability of a potential short-term sexual partner accepted by men were lower than those accepted by women, in terms of both the overall desirability and the desirability of individual traits (e.g., social status, attractiveness) [@kenrick_evolution_1990; @regan_minimum_1998]. Another study also found that when presented with identical descriptions of potential mates, men on average rated them as more desirable than women did [@wiederman_evolution_1998].

@@

Since heterosexual sex in most cases involves one man and one woman, men's relatively high interest in short-term mating can be constrained by women's preference for long-term relationships [@archer_reality_2019; @symons_evolution_1979]. This is because men’s short-term mating preferences – with the exception of sexual violence – can only translate into behaviors when there are women willing to have sex with them. When a new short-term mating encounter occurs, it counts towards both men’s total number of short-term mating encounters as well as women’s, and both men and women have new short-term mates. Therefore, the total number of short-term mating experiences and short-term mates must be equal between heterosexual men and women at the population level. In a closed population with equal sex ratio, it follows that the average number of experiences and mates should also be equal between men and women. Following this line of reasoning, we would expect to observe no gender difference in short-term mating behaviors among heterosexual individuals even when there were gender differences in mating preferences.

However, it is important to note that although heterosexual men’s total number of short-term matings would be equal to that of women, this does not necessarily mean that the total number of men and women who have ever had short-term mating must be equal. Empirical studies have found that the proportion of individuals who are sexually inexperienced is slightly higher among men than women in most age groups above 18 years old [@haydon2014prevalence; @ghaznavi2019trends]. Therefore, as a postulation, there may be a smaller proportion of men (vs. the proportion of women) who participate in the mating pool and contribute to the total number of short-term mating. For example, suppose that there is a population of 20 individuals with an equal sex ratio. Heterosexual short-term matings have happened \(n\) times in this population, but only 3 men and 8 women have had any short-term mating experiences. Since the total number of short-term mating behaviors is the same between men and women, we would expect that in a population of individuals who had any short-term mating, the average number of short-term mating experiences and mates was higher among men than women (e.g., in the aforementioned example, the average number of short-term experiences is \(\frac{n}{3}\) for men and \(\frac{n}{8}\) for women who are in the mating pool).

@@

The model measured two outcomes: (1) the number of short-term mating experiences of men and women; (2) the number of short-term mates of men and women. Additionally, we also measured the number of men and women in the mating pool (i.e., those with a non-zero number of short-term mating experiences). The average numbers of short-term mating experiences and short-term mates were calculated by taking the average among the whole population of men and women (*N~men~* = *N~women~* = 150) and among those who had engaged in any short-term mating.

@@

### Procedures
#### Procedures for heterosexual agents.
At each time step, agents first checked whether they were in a long-term relationship. If they were not in a long-term relationship, they set their heading randomly (if they were within the movement range from the birthplace) or faced the birthplace (if they were out of the movement range from the birthplace) and moved by a random distance. The random distance was less than half of the movement range. If they were in a long-term relationship, they did not move. Then, the agents decided on whether to engage in short-term mating using the predetermined likelihood (either 25% or 40%).

@@

Then, women randomly selected one man at the same location as their potential long-term partner. There was a 10% of chance that a pair would form a long-term relationship. After forming a long-term relationship, both men and women changed to coupled status and registered each other as their long-term partner.

@@

\begin{figure}[h]
  \centering
  \includegraphics[width=100mm]{figures/fig1_men_vs_women.pdf}
  \caption{\textbf{Short-term mating behaviors of men and women after 1,000 time steps in the model when gender differences existed in mating preferences.} Violin plots summarizing the outcome variables separately for heterosexual individuals and gay men and lesbian women. Plot (A) shows the average number of short-term mating experiences, and plot (B) shows the average number of short-term mates. Central points show mean values and whiskers represent standard errors, but the standard errors are small and overlap to form a single bar. The averages were calculated using the full population of men and women in the model (\textit{N\textsubscript{men}} = \textit{N\textsubscript{women}} = 150).}
  \label{fig:men_vs_women}
\end{figure}

@@

\begin{figure}[h]
  \centering
  \includegraphics[width=0.8\columnwidth]{figures/fig2_men_vs_women_inpool.pdf}
  \caption{\textbf{Short-term mating behaviors of heterosexual men and women in the mating pool after 1,000 time steps in the model when gender differences existed in mating preferences.} Violin plots summarizing the outcome variables. Plot (A) shows the average number of short-term mating experiences in the mating pool, and plot (B) shows the average number of short-term mates in the mating pool. Central points show mean values and whiskers represent standard errors, but the standard errors are small and overlap to form a single bar. The averages were calculated using the population of heterosexual men and women in the mating pool (proportion: \textit{M\textsubscript{men}} = `r m_inpool.mean_011`, \textit{M\textsubscript{women}} = `r f_inpool.mean_011`).}
  \label{fig:men_vs_women_inpool}
\end{figure}

@@

Gay men engaged in short-term mating behaviors more than heterosexual men did (Figure \ref{fig:hetero_vs_gay}). The average number of short-term mating experiences (as calculated using the full population of men) was higher among gay men (*M* = `r m_exp_all.mean_1_11`, *SD* = `r m_exp_all.sd_1_11`) than among heterosexual men (*M* = `r m_exp_all.mean_0_11`, *SD* = `r m_exp_all.sd_0_11`), *t*(`r exp.all.parameter_hg`) = `r exp.all.statistic_hg`, *p* `r exp.all.p.value_hg`, Cohen's *d* = `r exp_all.d_hg`, `r exp_all.ci_hg`. Similarly, the average number of short-term mates (as calculated using the full population of men) was higher among gay men (*M* = `r m_partner_all.mean_1_11`, *SD* = `r m_partner_all.sd_1_11`) than among heterosexual men (*M* = `r m_partner_all.mean_0_11`, *SD* = `r m_partner_all.sd_0_11`), *t*(`r partner.all.parameter_hg`) = `r partner.all.statistic_hg`, *p* `r partner.all.p.value_hg`, Cohen's *d* = `r partner_all.d_hg`, `r partner_all.ci_hg`.

\begin{figure}[h]
  \centering
  \includegraphics[width=0.8\columnwidth]{figures/fig3_hetero_vs_gay_men.pdf}
  \caption{\textbf{Short-term mating behaviors of heterosexual and gay men after 1,000 time steps in the model when gender differences existed in mating preferences.} Violin plots summarizing the outcome variables. Plot (A) shows the average number of short-term mating experiences, and plot (B) shows the average number of short-term mates. Central points show mean values and whiskers represent standard errors, but the standard errors are small and overlap to form a single bar. The averages were calculated using the full population of heterosexual and gay men in the model (\textit{N\textsubscript{hetero}} = \textit{N\textsubscript{gay}} = 150).}
  \label{fig:hetero_vs_gay}
\end{figure}

@@

```{r table_1, include=TRUE, echo=FALSE}
results %>%
  filter(hls %in% c("0_0_0", "0_0_1", "0_1_0", "0_1_1")) %>%
  select(m_exp_ip.mean, m_exp_ip.sd,
         f_exp_ip.mean, f_exp_ip.sd, exp_ip.d,
         m_partner_ip.mean, m_partner_ip.sd,
         f_partner_ip.mean, f_partner_ip.sd, partner_ip.d) %>%
  mutate(
    exp_ip.d = abs(exp_ip.d),
    partner_ip.d = abs(partner_ip.d)
  ) %>%
  mutate(across(where(is.numeric), inline_hook)) %>%
  mutate(
    condition =
      c("Same likelihood*Same standard",
        "Same likelihood*Different standard",
        "Different likelihood*Same standard",
        "Different likelihood*Different standard")
  ) %>%
  column_to_rownames("condition") %>%
  kable(
    digits = 2, format = "latex", booktabs = TRUE,
    caption = "\\textbf{Short-term mating behaviors of heterosexual men and women in the mating pool after 1,000 time steps in the model}",
    row.names = TRUE, col.names = c(rep(c(rep(c("M", "SD"), 2), "d"), 2))
    ) %>%
  row_spec(0, italic = TRUE, align = "c") %>%
  add_header_above(
    c(" " = 1,
      "Men" = 2, "Women" = 2, " " = 1,
      "Men" = 2, "Women" = 2, " " = 1)
    ) %>%
  add_header_above(
    c(" " = 1,
      "Average short-term mating\nexperiences (in-pool)" = 5,
      "Average short-term\nmates (in-pool)" = 5)
    ) %>%
  column_spec(1, width = "2in") %>%
  footnote(
    general = "\\\\textit{Note.} The averages were calculated with subsamples of men and women who had had engaged in short-term mating behaviors. See Supplemental Materials for the mean and standard deviation of sample sizes.",
    general_title = "", escape = FALSE,
    threeparttable = TRUE, fixed_small_size = TRUE) %>%
  kable_styling(latex_options = "scale_down")
```

@@

Among gay men and lesbian women, gender differences in short-term mating behaviors emerged both when they had different short-term mating likelihood and when their mating standards were different. Gay men (vs. lesbian women) had a higher average number of short-term mating experiences and short-term mates when they had a higher short-term mating likelihood (experiences: *M~men~* = `r m_exp_all.mean_110`, *SD~men~* = `r m_exp_all.sd_110`, *M~women~* = `r f_exp_all.mean_110`, *SD~women~* = `r f_exp_all.sd_110`, Cohen's *d* = `r exp_all.d_110`; partners: *M~men~* = `r m_partner_all.mean_110`, *SD~men~* = `r m_partner_all.sd_110`, *M~women~* = `r f_partner_all.mean_110`, *SD~women~* = `r f_partner_all.sd_110`, Cohen's *d* = `r partner_all.d_110`), or when they had lower mating standards (experiences: *M~men~* = `r m_exp_all.mean_101`, *SD~men~* = `r m_exp_all.sd_101`, *M~women~* = `r f_exp_all.mean_101`, *SD~women~* = `r f_exp_all.sd_101`, Cohen's *d* = `r exp_all.d_101`; partners: *M~men~* = `r m_partner_all.mean_101`, *SD~men~* = `r m_partner_all.sd_101`, *M~women~* = `r f_partner_all.mean_101`, *SD~women~* = `r f_partner_all.sd_101`, Cohen's *d* = `r partner_all.d_101`). In comparison, virtually no gender differences in short-term mating behaviors existed when gay men and lesbian women had the same short-term mating likelihood and mating standards (Table 2).

@@

The present study used agent-based modeling to investigate whether men and women’s differential preferences for short-term mating would result in gender differences in short-term mating behaviors, and if so, then under what conditions they would result in differences in the number of short-term mating experiences and short-term mates. We formulated five hypotheses: 1) there would be no gender differences in short-term mating behaviors between heterosexual men and women when considering the full population, 2) a smaller proportion of heterosexual men (vs. heterosexual women) would have ever engaged in short-term mating, 3) in a population of heterosexual individuals who had any short-term mating, men would have a higher average number of short-term mating behaviors than women, 4) gay men would engage in more short-term mating behaviors compared to lesbian women, and 5) compared to heterosexual men. The results from 1,000 time steps in a model simulating men and women’s mating behaviors provided strong evidence in favor of our hypotheses. First of all, we found no gender differences in short-term mating behaviors between heterosexual men and women when the full population was considered. However, compared to heterosexual women, there was a smaller proportion of heterosexual men participating in the mating pool, but they had higher average numbers of short-term mating experiences and short-term mates. Secondly, as compared to lesbian women, gay men had higher average numbers of short-term mating experiences and short-term mates. Finally, gay men also had higher average numbers of short-term mating experiences and short-term mates as compared to heterosexual men.

As expected, when the full population was considered, heterosexual men and women did not differ in short-term mating behaviors despite their differential preferences for short-term mating. This was because heterosexual men and women had an equal total number of both short-term mating experiences and short-term mates. This is not surprising since the sex ratio was 1:1 in our model. It therefore follows that the average number of short-term mating experiences and short-term mates must be equal between men and women as well. However, we did find that when we only looked at individuals in the mating pool (from which more men than women were excluded), men engaged in more short-term mating behaviors as compared to women. This was because there were more women than men in the mating pool, resulting in lower averages among heterosexual women despite the equal numbers of experiences and mates in total.

Moreover, gender differences in short-term mating behaviors emerged when heterosexual men and women had different mating standards, but not when they had different short-term mating likelihood. When women had a higher standard than men did, less men than women in the population were above a potential partner’s standard and thus had a chance to have sex with them. This contributed to the unequal number of men and women in the mating pool, which led to gender differences in short-term mating behaviors among this population. In comparison, even when men and women had different short-term mating likelihood, the probability of any pair of them ending up having sex is the product of their individual likelihood (since men and women made decisions independently). As this probability is the same for both parties, the gender difference in short-term mating likelihood did not translate into different behaviors.

@@

Among gay men and lesbian women, large gender differences in short-term mating behaviors existed when men and women had differential preferences for short-term mating, which was in line with empirical observations [e.g., @peplau_national_1997; @peplau_sexuality_2004]. This was likely because the number of short-term mating experiences and short-term mates no longer counted towards men and women simultaneously. Any gender differences in mating preferences would result in differences in behaviors. A closer look at the results did support this postulation. Both a difference in the likelihood of short-term mating and in mating standards alone resulted in gender differences in mating behaviors. This was likely because the former increased the probability of both parties of a given gay couple deciding to have short-term mating, and the latter increased the probability of both meeting each other's standards, as compared to the case of a given lesbian couple.

@@

Using agent-based modeling to simulate short-term mating behaviors, we found when men (vs. women) had stronger preferences for short-term mating, heterosexual men and women engaged in short-term mating behaviors to the same extent when the full population was considered, while men in the mating pool had more such behaviors than women in the mating pool. Gay men engaged in more short-term mating behaviors compared to both lesbian women and heterosexual men.

These results highlight the distinction between preferences and behaviors in human mating. Individuals’ mating behaviors do not only depend on one’s own preferences, but are also constrained by partners’ preferences. Future research in human mating should not only focus on the psychological aspect but also pay attention to the interaction between individuals’ psychology and its behavioral context. These results also cast doubt to the prevalent belief in the gender differences in short-term mating behaviors, especially among heterosexual individuals. Our findings suggest that there may be factors in the observation process, such as sampling bias, that contribute to the observed differences. Future research in human sexuality should note such possibilities and interpret any observed gender differences in short-term mating behaviors with caution.

@@

We thank Gu Li and Jiannan Shi for providing helpful feedback in the development of this study.

@@

All models, data, and analysis code can be downloaded at: [link masked for peer review]

>>>>

@@

@@

Gender differences in short-term mating behaviors are well-documented in human sexuality research. Existing studies usually conflate gender differences in mating preferences with differences in sexual behaviors, which is theoretically problematic. Using an agent-based model, we investigated the circumstances under which men and women’s differential preferences for short-term mating would result in gender differences in short-term mating behaviors. The model showed that when all individuals in a closed heterosexual population were considered,  men and women had the same average number of short-term mating experiences and short-term mates even when men had stronger preferences for short-term mating. Men (vs. women) had a higher average number of both experiences and mates when analyses were limited to only heterosexual men and women who successfully participated in the mating pool (i.e., those with a non-zero number of short-term mating experiences). Moreover, when men (vs. women) had stronger preferences for short-term mating, gay men had a higher average number of experiences and mates compared to both lesbian women and heterosexual men. These results suggest that even when gender differences in mating preferences exist, those in short-term mating behaviors only occur among particular populations, or when men’s preferences for short-term mating are not constrained by those of women. Suggestions for future research in human mating psychology and behaviors were provided.

@@

As for mating standards, men are less selective in accepting someone as a potential short-term mate. For example, studies using U.S. college samples have found that men' acceptable minimum percentile ranks of the desirability of a potential short-term sexual partner were lower than those accepted by women, in terms of both the overall desirability and the desirability of individual traits (e.g., social status, attractiveness) [@kenrick_evolution_1990; @regan_minimum_1998]. Another study also found that when presented with identical descriptions of potential mates, men on average rated them as more desirable than women did [@wiederman_evolution_1998].

@@

Since heterosexual sex in most cases involves one man and one woman, men's relatively high interest in short-term mating can be constrained by women's preference for long-term relationships [@archer_reality_2019; @symons_evolution_1979]. This is because men’s short-term mating preferences – with the exception of sexual violence – can only translate into behaviors when there are women willing to have sex with them. When a new short-term mating encounter occurs, both men’s and women's total number of short-term mating experiences and the number of short-term mates increase by one. Therefore, the total number of short-term mating experiences and short-term mates must be equal between heterosexual men and women at the population level. In a closed population with equal sex ratio, it follows that the average number of experiences and mates should also be equal between men and women. Following this line of reasoning, we would expect to observe no gender difference in short-term mating behaviors among heterosexual individuals even when there were gender differences in mating preferences.

However, it is important to note that although heterosexual men’s total number of short-term matings would be equal to that of women, this does not necessarily mean that the total number of men and women who have ever had short-term mating must be equal. Empirical studies have found that the proportion of individuals who are sexually inexperienced is slightly higher among men than women in most age groups above 18 years old [@haydon2014prevalence; @ghaznavi2019trends]. Therefore, as a postulation, there may be a smaller proportion of men (vs. the proportion of women) who participate in the mating pool and contribute to the total number of short-term mating. For example, suppose that there is a population of 20 individuals with an equal sex ratio. Heterosexual short-term matings have happened \(n\) times in this population, but only 3 men and 8 women have had any short-term mating experiences. Since the total number of short-term mating behaviors is the same between men and women, we would expect that in a population of individuals who participate in the mating pool, the average numbers of short-term mating experiences and mates are higher among men than women (e.g., in the aforementioned example, the average number of short-term experiences is \(\frac{n}{3}\) for men and \(\frac{n}{8}\) for women who are in the mating pool).

@@

The model measured two outcomes: (1) the number of short-term mating experiences of men and women; (2) the number of short-term mates of men and women. Additionally, we also measured the number of men and women in the mating pool (i.e., those with a non-zero number of short-term mating experiences). The average numbers of short-term mating experiences and short-term mates were calculated by taking the average among the whole population of men and women (*N~men~* = *N~women~* = 150) and among those who had ever engaged in any short-term mating.

@@

### Procedures
#### Procedures for heterosexual agents.
At each time step, agents first checked whether they were in a long-term relationship. If they were in a long-term relationship, they did not move. If they were not in a long-term relationship, they set their heading randomly (if they were within the movement range from the birthplace) or faced the birthplace (if they were out of the movement range from the birthplace) and moved by a random distance. The random distance was less than half of the movement range. Then, the agents decided on whether to engage in short-term mating using the predetermined likelihood (either 25% or 40%).

@@

Then, women randomly selected one man at the same location as their potential long-term partner. There was a 10% of chance that a pair of individuals would form a long-term relationship. After forming a long-term relationship, both men and women changed to coupled status and registered each other as their long-term partner.

@@

\begin{figure}[h]
  \centering
  \includegraphics[width=100mm]{figures/fig1_men_vs_women.pdf}
  \caption{\textbf{Short-term mating behaviors of men and women after 1,000 time steps in the model when gender differences existed in mating preferences.} Violin plots summarizing the outcome variables separately for heterosexual individuals and gay men and lesbian women. Plot (A) shows the average number of short-term mating experiences, and plot (B) shows the average number of short-term mates. Central points show mean values and whiskers represent standard errors, but the standard errors are small and overlap to form a single bar. The means were calculated using the full population of men and women in the model (\textit{N\textsubscript{men}} = \textit{N\textsubscript{women}} = 150).}
  \label{fig:men_vs_women}
\end{figure}

@@

\begin{figure}[h]
  \centering
  \includegraphics[width=0.8\columnwidth]{figures/fig2_men_vs_women_inpool.pdf}
  \caption{\textbf{Short-term mating behaviors of heterosexual men and women in the mating pool after 1,000 time steps in the model when gender differences existed in mating preferences.} Violin plots summarizing the outcome variables. Plot (A) shows the average number of short-term mating experiences in the mating pool, and plot (B) shows the average number of short-term mates in the mating pool. Central points show mean values and whiskers represent standard errors, but the standard errors are small and overlap to form a single bar. The means were calculated using the population of heterosexual men and women who successfully participated in the mating pool (proportion: \textit{M\textsubscript{men}} = `r m_inpool.mean_011`, \textit{M\textsubscript{women}} = `r f_inpool.mean_011`).}
  \label{fig:men_vs_women_inpool}
\end{figure}

@@

Gay men engaged in short-term mating behaviors more than heterosexual men did (Figure \ref{fig:hetero_vs_gay}). The average number of short-term mating experiences (calculated using the full population of men) was higher among gay men (*M* = `r m_exp_all.mean_1_11`, *SD* = `r m_exp_all.sd_1_11`) than among heterosexual men (*M* = `r m_exp_all.mean_0_11`, *SD* = `r m_exp_all.sd_0_11`), *t*(`r exp.all.parameter_hg`) = `r exp.all.statistic_hg`, *p* `r exp.all.p.value_hg`, Cohen's *d* = `r exp_all.d_hg`, `r exp_all.ci_hg`. Similarly, the average number of short-term mates (calculated using the full population of men) was higher among gay men (*M* = `r m_partner_all.mean_1_11`, *SD* = `r m_partner_all.sd_1_11`) than among heterosexual men (*M* = `r m_partner_all.mean_0_11`, *SD* = `r m_partner_all.sd_0_11`), *t*(`r partner.all.parameter_hg`) = `r partner.all.statistic_hg`, *p* `r partner.all.p.value_hg`, Cohen's *d* = `r partner_all.d_hg`, `r partner_all.ci_hg`.

\begin{figure}[h]
  \centering
  \includegraphics[width=0.8\columnwidth]{figures/fig3_hetero_vs_gay_men.pdf}
  \caption{\textbf{Short-term mating behaviors of heterosexual men and gay men after 1,000 time steps in the model when gender differences existed in mating preferences.} Violin plots summarizing the outcome variables. Plot (A) shows the average number of short-term mating experiences, and plot (B) shows the average number of short-term mates. Central points show mean values and whiskers represent standard errors, but the standard errors are small and overlap to form a single bar. The means were calculated using the full population of heterosexual men and gay men in the model (\textit{N\textsubscript{hetero}} = \textit{N\textsubscript{gay}} = 150).}
  \label{fig:hetero_vs_gay}
\end{figure}

@@

```{r table_1, include=TRUE, echo=FALSE}
results %>%
  filter(hls %in% c("0_0_0", "0_0_1", "0_1_0", "0_1_1")) %>%
  select(m_exp_ip.mean, m_exp_ip.sd,
         f_exp_ip.mean, f_exp_ip.sd, exp_ip.d,
         m_partner_ip.mean, m_partner_ip.sd,
         f_partner_ip.mean, f_partner_ip.sd, partner_ip.d) %>%
  mutate(
    exp_ip.d = abs(exp_ip.d),
    partner_ip.d = abs(partner_ip.d)
  ) %>%
  mutate(across(where(is.numeric), inline_hook)) %>%
  mutate(
    condition =
      c("Same likelihood*Same standard",
        "Same likelihood*Different standard",
        "Different likelihood*Same standard",
        "Different likelihood*Different standard")
  ) %>%
  column_to_rownames("condition") %>%
  kable(
    digits = 2, format = "latex", booktabs = TRUE,
    caption = "\\textbf{Short-term mating behaviors of heterosexual men and women in the mating pool after 1,000 time steps in the model}",
    row.names = TRUE, col.names = c(rep(c(rep(c("M", "SD"), 2), "d"), 2))
    ) %>%
  row_spec(0, italic = TRUE, align = "c") %>%
  add_header_above(
    c(" " = 1,
      "Men" = 2, "Women" = 2, " " = 1,
      "Men" = 2, "Women" = 2, " " = 1)
    ) %>%
  add_header_above(
    c(" " = 1,
      "Average short-term mating\nexperiences (in-pool)" = 5,
      "Average short-term\nmates (in-pool)" = 5)
    ) %>%
  column_spec(1, width = "2in") %>%
  footnote(
    general = "\\\\textit{Note.} The means were calculated with subsamples of men and women who had ever engaged in short-term mating behaviors. See Supplemental Materials for the mean and standard deviation of sample sizes.",
    general_title = "", escape = FALSE,
    threeparttable = TRUE, fixed_small_size = TRUE) %>%
  kable_styling(latex_options = "scale_down")
```

@@

Among gay men and lesbian women, gender differences in short-term mating behaviors emerged both when they had different short-term mating likelihood and when they had different mating standards. Gay men (vs. lesbian women) had a higher average number of short-term mating experiences and short-term mates when they had a higher short-term mating likelihood (experiences: *M~men~* = `r m_exp_all.mean_110`, *SD~men~* = `r m_exp_all.sd_110`, *M~women~* = `r f_exp_all.mean_110`, *SD~women~* = `r f_exp_all.sd_110`, Cohen's *d* = `r exp_all.d_110`; partners: *M~men~* = `r m_partner_all.mean_110`, *SD~men~* = `r m_partner_all.sd_110`, *M~women~* = `r f_partner_all.mean_110`, *SD~women~* = `r f_partner_all.sd_110`, Cohen's *d* = `r partner_all.d_110`), or when they had lower mating standards (experiences: *M~men~* = `r m_exp_all.mean_101`, *SD~men~* = `r m_exp_all.sd_101`, *M~women~* = `r f_exp_all.mean_101`, *SD~women~* = `r f_exp_all.sd_101`, Cohen's *d* = `r exp_all.d_101`; partners: *M~men~* = `r m_partner_all.mean_101`, *SD~men~* = `r m_partner_all.sd_101`, *M~women~* = `r f_partner_all.mean_101`, *SD~women~* = `r f_partner_all.sd_101`, Cohen's *d* = `r partner_all.d_101`). In comparison, virtually no gender differences in short-term mating behaviors existed when gay men and lesbian women had the same short-term mating likelihood and mating standards (Table 2).

@@

The present study used agent-based modeling to investigate whether men and women’s differential preferences for short-term mating would result in gender differences in short-term mating behaviors, and if so, then under what conditions they would result in differences in the number of short-term mating experiences and short-term mates. We formulated five hypotheses: 1) there would be no gender differences in short-term mating behaviors between heterosexual men and women when considering the full population, 2) a smaller proportion of heterosexual men (vs. heterosexual women) would have ever engaged in short-term mating, 3) in a population of heterosexual individuals who had any short-term mating, men would have a higher average number of short-term mating behaviors than women, 4) gay men would engage in more short-term mating behaviors compared to lesbian women, and 5) compared to heterosexual men. The results from 1,000 time steps in a model simulating men and women’s mating behaviors provided strong evidence in favor of our hypotheses. First of all, we found no gender differences in short-term mating behaviors between heterosexual men and women when the full population was considered. However, compared to heterosexual women, there was a smaller proportion of heterosexual men participating in the mating pool, but they had higher average numbers of short-term mating experiences and short-term mates. Secondly, gay men had higher average numbers of short-term mating experiences and short-term mate compared to both lesbian women and heterosexual men.

As expected, when all individuals in the population were considered, heterosexual men and women did not differ in short-term mating behaviors despite their differential preferences for short-term mating. This is not surprising since the sex ratio was 1:1 in our model. Heterosexual men and women had an equal total number of both short-term mating experiences and short-term mates. It therefore follows that the average number of short-term mating experiences and short-term mates must be equal between men and women as well. However, we did find that when we only looked at individuals in the mating pool (from which more men than women were excluded), men engaged in more short-term mating behaviors compared to women. This was because there were more women than men in the mating pool, resulting in lower averages among heterosexual women despite the equal numbers of experiences and mates in total.

Moreover, gender differences in short-term mating behaviors emerged among individuals in the mating pool when heterosexual men and women had different mating standards, but not when they had different short-term mating likelihood. When women had a higher standard than men, less men than women in the population were above a potential partner’s standard and thus had a chance to have sex with them. This contributed to the unequal number of men and women in the mating pool, which led to gender differences in short-term mating behaviors among this population. In comparison, when men and women had different short-term mating likelihood, the probability of any pair of them ending up having sex is always the product of their individual likelihood (since men and women made decisions independently). As this probability is the same for both parties, the gender difference in short-term mating likelihood did not translate into different behaviors.

@@

Among gay men and lesbian women, large gender differences in short-term mating behaviors existed when men and women had differential preferences for short-term mating, which was in line with empirical observations [e.g., @peplau_national_1997; @peplau_sexuality_2004]. This was likely because the number of short-term mating experiences and short-term mates no longer counted towards men and women simultaneously. Any gender differences in mating preferences would result in differences in behaviors. A closer look at the results did support this postulation. Both a difference in the likelihood of short-term mating and in mating standards alone resulted in gender differences in mating behaviors. This was likely because the former increased the probability of both parties of a given gay couple deciding to have short-term mating, and the latter increased the probability of both meeting each other's standards, compared to the case of a given lesbian couple.

@@

Using agent-based modeling to simulate short-term mating behaviors, we found when men (vs. women) had stronger preferences for short-term mating, heterosexual men and women engaged in short-term mating behaviors to the same extent when all individuals in the population were considered. However, among individuals who participated in the mating pool, men had more such behaviors than women. Gay men engaged in more short-term mating behaviors compared to both lesbian women and heterosexual men.

These results highlight the distinction between preferences and behaviors in human mating. Individuals’ mating behaviors do not only depend on one’s own preferences, but are also constrained by partners’ preferences. Future research should not only focus on the psychological aspect of human mating but also pay attention to the interaction between individuals’ psychology and its behavioral context. These results also cast doubt to the prevalent belief in the gender differences in short-term mating behaviors, especially among heterosexual individuals. Our findings suggest that there may be factors in the observation process, such as sampling bias, that contribute to the observed differences. Future research in human sexuality should note such possibilities and interpret any observed gender differences in short-term mating behaviors with caution.

@@

We thank [name masked for peer review] and [name masked for peer review] for providing helpful feedback in the development of this study. This research did not receive any specific grant from funding agencies in the public, commercial, or not-for-profit sectors.

@@

All model, data, and analysis code (including its history) are available at: [link masked for peer review].
